---
title: "Fastest way to calculate BED12 blocks for UCSCData objects"
author: "Charles Plessy"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

In the [BED12 format](https://genome.ucsc.edu/FAQ/FAQformat.html#format1),
the position of each box in the object is encoded in the `blockCount`,
`blockSize` and `blockStarts` columns.

In `UCSCData` track objects of the _rtracklayer_ package, the information is
encoded as an `IRangesList` metadata column, where each element has one
`IRanges` object representing the blocks.

In _CAGEr_'s `TagClusters` objects, the position of the central blocks is
represented by the quantile information columns.

The problem here is to create this `IRanges` list in a computationally efficient
way in _CAGEr_'s `exportToTrack` function.


The best approach, `f.mat.direct` is not exactly fast, but is 5 × faster than my
original attempt (`f.grline`).

```{r setup, echo=FALSE}
options(width=120)
knitr::opts_chunk$set(cache  = TRUE, cache.lazy = FALSE)
knitr::opts_knit$set(verbose = TRUE)
```

```{r create_data}
suppressPackageStartupMessages(library("GenomicRanges"))
suppressPackageStartupMessages(library("CAGEr"))
tc <- CAGEr::exampleCAGEexp |> CAGEr::tagClustersGR(1)
tcl <- split(tc, seq_along(tc))
decoded.mat <- rbind(width(tc), decode(tc$q_0.1), decode(tc$q_0.9))
```

```{r declare_functions}

# Operate directly on each line

f.grline <- function(grline) {
  # A GRanges line is a GRanges object of length 1
  # This function computes a blocks value for each line
  qLow_value <- decode(grline$q_0.1)
  qUp_value  <- decode(grline$q_0.9)
  ir <- IRanges()                      |>
    c( if(qLow_value != 1) IRanges(1)) |>
    c( IRanges(qLow_value, qUp_value)) |>
    c( if(qUp_value != width(grline)) IRanges(width(grline)))
  ir
}

# Decode the Rle objects once for all and loop on each entry

f.mat.direct <- function(x) {
  width_value <- x[1]
  qLow_value <- x[2]
  qUp_value  <- x[3]
  ir <- IRanges()                      |>
    c( if(qLow_value != 1) IRanges(1)) |>
    c( IRanges(qLow_value, qUp_value)) |>
    c( if(qUp_value != width_value) IRanges(width_value))
}

# Same but try to call the IRanges constructor only once.

f.mat.viaString <- function(x) {
  width_value <- x[1]
  qLow_value <- x[2]
  qUp_value  <- x[3]
  str <- c(
    c( if(qLow_value != 1) "1"),
    c( paste0(qLow_value, "-", qUp_value)),
    c( if(qUp_value != width_value) width_value)
  )
  IRanges(str)
}
```

```{r benchmark}
tc[1]
microbenchmark::microbenchmark(f.grline(tcl[[1]]), f.mat.direct(decoded.mat[,1]), f.mat.viaString(decoded.mat[,1]), times = 100)

tc[5]
microbenchmark::microbenchmark(f.grline(tcl[[5]]), f.mat.direct(decoded.mat[,5]), f.mat.viaString(decoded.mat[,5]), times = 100)
```
