---
title: "Cumulative sums"
author: "Charles Plessy"
date: "2023-07-18"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

## Purpose

The function to compute cumulative sums is one of the slowest in CAGEr.
Can we speed it up?

## Setup

```{r setup}
library("CAGEr")   |> suppressPackageStartupMessages()
library("ggplot2") |> suppressPackageStartupMessages()
```

### Example data

```{r define_data}
(clusters <- tagClustersGR(exampleCAGEexp)[[1]])
(ctss <- CTSSnormalizedTpmGR(exampleCAGEexp, 1))
```

## Current implementation at the time the benchmark was written.

```{r current_function}
cumsums_viewApply <- function(ctss, clusters) {
  cov <- Rle(rep(0, max(end(clusters), end(ctss))))
  cov[start(ctss)] <- score(ctss)
  cluster.cumsums <- Views(cov, start = start(clusters), end = end(clusters))
  viewApply(cluster.cumsums, cumsum)
}
cumsums_viewApply_res <- cumsums_viewApply(ctss, clusters)
head(cumsums_viewApply_res)
```

## Alternative functions.

`cumsums_extractList_sapply` is a reduced version of the `bioC2_cc_iqw` currently
used in `consensusClustersGR`.

```{r define_functions}
cumsums_extractList_sapply <- function(ctss, clusters) {
  # Fill gaps with zeros
  fill <- GPos(clusters)
  score(fill) <- Rle(0L)
  ctss <- c(ctss, fill[!fill %in% ctss]) |> sort()
  o <- findOverlaps(query = clusters, subject = ctss)
  grouped_scores <- extractList(score(ctss), o)
  sapply(grouped_scores, cumsum)
}
cumsums_extractList_sapply_res <- cumsums_extractList_sapply(ctss, clusters)

cumsums_extractList_aslist <- function(ctss, clusters) {
  # Fill gaps with zeros
  fill <- GPos(clusters)
  score(fill) <- Rle(0L)
  ctss <- c(ctss, fill[!fill %in% ctss]) |> sort()
  o <- findOverlaps(query = clusters, subject = ctss)
  grouped_scores <- extractList(score(ctss), o)
  cumsum(grouped_scores) |> as.list()
}
cumsums_extractList_aslist_res <- cumsums_extractList_aslist(ctss, clusters)

cumsums_extractList_RleList <- function(ctss, clusters) {
  # Fill gaps with zeros
  fill <- GPos(clusters)
  score(fill) <- Rle(0L)
  ctss <- c(ctss, fill[!fill %in% ctss]) |> sort()
  o <- findOverlaps(query = clusters, subject = ctss)
  grouped_scores <- extractList(score(ctss), o)
  cumsum(grouped_scores)
}
cumsums_extractList_RleList_res <- cumsums_extractList_RleList(ctss, clusters)
```

The results are almost the same.  The differences appear to be rounding errors.

```{r compare_results}
# Identical
identical(cumsums_viewApply_res[[3]],  cumsums_extractList_sapply_res[[3]])
cumsums_viewApply_res[[3]] - cumsums_extractList_sapply_res[[3]]
identical(cumsums_viewApply_res,  cumsums_extractList_sapply_res)

# Slightly different
identical(cumsums_viewApply_res[[3]], cumsums_extractList_RleList_res[[3]])
cumsums_viewApply_res[[3]] - cumsums_extractList_aslist_res[[3]]

# Identical
identical(cumsums_extractList_RleList_res[[3]], cumsums_extractList_aslist_res[[3]])
```

## Benchmark

```{r benchmark, fig.height=3, fig.width=9}
(microbench_out <- microbenchmark::microbenchmark(
  times = 100,
  viewApply = cumsums_viewApply(ctss, clusters),
  extractList_sapply  = cumsums_extractList_sapply(ctss, clusters),
  extractList_lapply  = cumsums_extractList_aslist(ctss, clusters),
  extractList_RleList = cumsums_extractList_RleList(ctss, clusters)
))

# https://statisticsglobe.com/microbenchmark-package-r
ggplot(microbench_out, aes(x = time / 1e6, y = expr, color = expr)) +  # Plot performance comparison
  geom_boxplot() + 
  scale_x_log10("time (milliseconds)")
```

## Result

The winner is about 10 times faster!

```{r}
sessionInfo()
```