---
title: "Paraclu benchmark"
author: "Charles Plessy"
date: "2023-07-14"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(CAGEr) |> suppressWarnings()
```

## Purpose

While refactoring the Paraclu code, I wondered if some data structures would be
more efficient than `data.frame`s.

## Example data

Let's focus on the clustering of positions with scores from one strand of one
chromosome, because this is where the algorithm is recursive and hard to
parallelise.

In the original implementation the input in two columns of a `data.frame`.  Here
I also try a `Pairs` object, or just two separate arguments for the positions
and scores.

```{r example_data}
ctss <- CTSSnormalizedTpmGR(exampleCAGEexp,1)
isSorted(ctss)
ctss.df <- as.data.frame(granges(ctss))
ctss.df$tpm <- decode(score(ctss))
ctss.df$pos <- ctss.df$start
ctss.df$end <- ctss.df$start <- ctss.df$width <- NULL

pair <- Pairs(pos(ctss), decode(score(ctss)))
DF <- DataFrame(pos = pos(ctss), score = decode(score(ctss)))
l <- list(pos = pos(ctss), score = decode(score(ctss)))
```

## The paraclu parameters function

The main recursive functions call a subfunction that compute the density and
the next break point.  First, let's see if it can be optimised.

`paraclu_params_df` was originally called `.paraclu1`.  The other functions
are rewrites.

```{r paraclu_params}
paraclu_params_df <- function(ctss) {
	sit <- nrow(ctss)
	tot <- sum(ctss$tpm)
	if(sit == 1) {
		min_density <- Inf
		br <- NA
	}else{
		densities_forward <- cumsum(ctss$tpm)[-sit]/(ctss$pos[2:sit] - ctss$pos[1])
		densities_reverse <- cumsum(rev(ctss$tpm))[-sit]/(ctss$pos[sit] - ctss$pos[(sit-1):1])
		min_densities = c(min(densities_forward), min(densities_reverse))
		breaks <- c(which(densities_forward == min_densities[1])[1] + 1, sit + 1 - which(densities_reverse == min_densities[2])[1])
		min_density <- min(min_densities)
		br <- breaks[tail(which(min_densities == min_density),1)]
	}
	list(br = br, min_density = min_density, tot = tot, sit = sit)
}
paraclu_params_df(ctss.df)

paraclu_params_Pairs <- function(pair) {
  sit   <- length(pair)
  score <- second(pair)
  tot   <- sum(score)
  if(sit == 1) return(list(br = NA, min_density = Inf, tot = tot, sit = sit))
  pos   <- first(pair)
  densities_forward <- cumsum(    score) [-sit] / (pos[2:sit] - pos[1]        )
  densities_reverse <- cumsum(rev(score))[-sit] / (pos[sit]   - pos[(sit-1):1])
  min_densities = c(min(densities_forward), min(densities_reverse))
  breaks <- c(     1 +   which(densities_forward == min_densities[1])[1]
             , sit + 1 - which(densities_reverse == min_densities[2])[1])
  min_density <- min(min_densities)
  br <- breaks[tail(which(min_densities == min_density),1)]
  list(br = br, min_density = min_density, tot = tot, sit = sit)
}
identical(paraclu_params_Pairs(pair), paraclu_params_df(ctss.df))

paraclu_params_DF <- function(DF) {
  sit   <- nrow(DF)
  tot   <- sum(DF$score)
  if(sit == 1) return(list(br = NA, min_density = Inf, tot = tot, sit = sit))
  densities_forward <- cumsum(    DF$score) [-sit] / (DF$pos[2:sit] - DF$pos[1]        )
  densities_reverse <- cumsum(rev(DF$score))[-sit] / (DF$pos[sit]   - DF$pos[(sit-1):1])
  min_densities = c(min(densities_forward), min(densities_reverse))
  breaks <- c(     1 +   which(densities_forward == min_densities[1])[1]
             , sit + 1 - which(densities_reverse == min_densities[2])[1])
  min_density <- min(min_densities)
  br <- breaks[tail(which(min_densities == min_density),1)]
  list(br = br, min_density = min_density, tot = tot, sit = sit)
}
identical(paraclu_params_DF(DF), paraclu_params_df(ctss.df))

paraclu_params_twoargs <- function(pos, score) {
  sit   <- length(pos)
  tot   <- sum(score)
  if(sit == 1) return(list(br = NA, min_density = Inf, tot = tot, sit = sit))
  densities_forward <- cumsum(    score) [-sit] / (pos[2:sit] - pos[1]        )
  densities_reverse <- cumsum(rev(score))[-sit] / (pos[sit]   - pos[(sit-1):1])
  min_densities = c(min(densities_forward), min(densities_reverse))
  breaks <- c(     1 +   which(densities_forward == min_densities[1])[1]
             , sit + 1 - which(densities_reverse == min_densities[2])[1])
  min_density <- min(min_densities)
  br <- breaks[tail(which(min_densities == min_density),1)]
  list(br = br, min_density = min_density, tot = tot, sit = sit)
}
identical(paraclu_params_twoargs(l$pos, l$score), paraclu_params_df(ctss.df))

microbenchmark::microbenchmark(
  data.frame = paraclu_params_df(ctss.df),
  Pairs      = paraclu_params_Pairs(pair),
  DataFrame  = paraclu_params_DF(DF),
  two_args   = paraclu_params_twoargs(l$pos, l$score),
  times = 1000)
```

All functions are microsecond-fast except the DF one.  Let's use the two-
arguments version in all the benchmarks below.

## Recursive Paraclu function

The original recursive function was originally called `.paraclu2`.  It computes
extra information like dominant peak etc, which is nice but [we have a good
function to do that _a posteriori_](.dominant_ctss.md).  Therefore we will compare
implementations that skip that calculation.

```{r paraclu2}
# Originally called .paraclu2
# Minor changes applied to keep ctss.df colnames.
# Removed re-sorting of output
paraclu_orig <- function(ctss, min_density = -Inf, clusters.df = data.frame()) {
	if(nrow(ctss)>0){
		params <- paraclu_params_twoargs(ctss$pos, ctss$tpm)
		br <- params[[1]]
		max_density <- params[[2]]
		tot <- params[[3]]
		sit <- params[[4]]
		
		if(!(max_density == Inf)){
			new_min <- max(min_density, max_density)
			clusters.df <- rbind(paraclu_orig(ctss = ctss[1:(br-1),], min_density = new_min, clusters.df = clusters.df), 
			                     paraclu_orig(ctss = ctss[br:nrow(ctss),], min_density = new_min, clusters.df = clusters.df))
		}
		
		return(rbind(clusters.df, data.frame(seqnames = ctss$seqnames[1], 
                    start = min(ctss$pos), 
                    end = max(ctss$pos), 
                    strand = ctss$strand[1], 
                    nr_ctss = sit, 
		                dominant_ctss = ctss$pos[which(ctss$tpm == max(ctss$tpm))[ceiling(length(which(ctss$tpm == max(ctss$tpm)))/2)]], 
		                tpm = tot, 
		                tpm.dominant_ctss = ctss$tpm[which(ctss$tpm == max(ctss$tpm))[ceiling(length(which(ctss$tpm == max(ctss$tpm)))/2)]],
		                min_d = min_density, max_d= max_density)
		             )
		       )
	}else{
		return(clusters.df)
	}
}
paraclu_orig(ctss.df) |> head()

paraclu_orig_simpler_output <- function(ctss, min_density = -Inf, clusters.df = data.frame()) {
	if(nrow(ctss)>0){
		params <- paraclu_params_twoargs(ctss$pos, ctss$tpm)
		br <- params[[1]]
		max_density <- params[[2]]
		tot <- params[[3]]
		sit <- params[[4]]
		
		if(!(max_density == Inf)){
			new_min <- max(min_density, max_density)
			clusters.df <- rbind(paraclu_orig_simpler_output(ctss = ctss[1:(br-1),], min_density = new_min, clusters.df = clusters.df), 
			                     paraclu_orig_simpler_output(ctss = ctss[br:nrow(ctss),], min_density = new_min, clusters.df = clusters.df))
		}
		
		return(rbind(clusters.df, data.frame(seqnames = ctss$seqnames[1], 
                    start = min(ctss$pos), 
                    end = max(ctss$pos), 
		                min_d = min_density, max_d= max_density)
		             )
		       )
	}else{
		return(clusters.df)
	}
}
paraclu_orig_simpler_output(ctss.df) |> head()
```

Representing the clusters in the `IRanges` class looks elegant.  But is it
efficient?  Also, since we need only two vectors, how about keeping them
together as a pair?

```{r paraclu_Pair_IRanges}
paraclu_Pair_IRanges <- function(pair, min_density = -Inf, clusters = IRanges()) {
  params <- paraclu_params_twoargs(first(pair), second(pair))

  if (!is.na(params$br)) {
    new_min     <- max(min_density, params$min_density)
    clusters    <- c(paraclu_Pair_IRanges(pair[   1      : (params$br-1)], new_min, clusters), 
                     paraclu_Pair_IRanges(pair[params$br :  length(pair)], new_min, clusters))
  }
  
  c( clusters
   , IRanges( start = min(first(pair))
            ,   end = max(first(pair))
            , min_d = min_density
            , max_d = params$min_density))
}
paraclu_Pair_IRanges(pair)
```

If it is too slow, is it because of the `Pairs` input?  Let's replace it with
two arguments for position and score separately.

```{r paraclu_on_Pairs}
paraclu_twoargs_IRanges <- function(pos, score, min_density = -Inf, clusters = IRanges()) {
  params <- paraclu_params_twoargs(pos, score)

  if (!is.na(params$br)) {
    new_min     <- max(min_density, params$min_density)
    left  <-    1      : (params$br-1)
    right <- params$br :  length(pos)
    clusters <- c(paraclu_twoargs_IRanges(pos[left],  score[left],  new_min, clusters), 
                  paraclu_twoargs_IRanges(pos[right], score[right], new_min, clusters))
  }
  
  c( clusters
   , IRanges( start = min(pos)
            ,   end = max(pos)
            , min_d = min_density
            , max_d = params$min_density))
}
paraclu_twoargs_IRanges(l$pos, l$score)
```

If it is still two slow, would it be better with a `DataFrame` output?

```{r paraclu_twoargs_DF}
paraclu_twoargs_DF <- function(pos, score, min_density = -Inf
                               , clusters = DataFrame()) {
  params <- paraclu_params_twoargs(pos, score)

  if (!is.na(params$br)) {
    new_min     <- max(min_density, params$min_density)
    left  <-    1      : (params$br-1)
    right <- params$br :  length(pos)
    clusters <- rbind(paraclu_twoargs_DF(pos[left],  score[left],  new_min, clusters), 
                      paraclu_twoargs_DF(pos[right], score[right], new_min, clusters))
  }
  
  rbind( clusters
   , DataFrame( start = min(pos)
            ,   end = max(pos)
            , min_d = min_density
            , max_d = params$min_density))
}
paraclu_twoargs_DF(l$pos, l$score)
```

Or a `tibble`?

```{r paraclu_twoargs_tibble}
paraclu_twoargs_tibble <- function(pos, score, min_density = -Inf
                               , clusters = tibble::tibble()) {
  params <- paraclu_params_twoargs(pos, score)

  if (!is.na(params$br)) {
    new_min     <- max(min_density, params$min_density)
    left  <-    1      : (params$br-1)
    right <- params$br :  length(pos)
    clusters <- rbind(paraclu_twoargs_tibble(pos[left],  score[left],  new_min, clusters), 
                      paraclu_twoargs_tibble(pos[right], score[right], new_min, clusters))
  }
  
  rbind( clusters
   , tibble::tibble( start = min(pos)
            ,   end = max(pos)
            , min_d = min_density
            , max_d = params$min_density))
}
paraclu_twoargs_tibble(l$pos, l$score)
```

Or a plain `data.frame`?

```{r paraclu_twoargs_df}
paraclu_twoargs_df <- function(pos, score, min_density = -Inf
                               , clusters = data.frame()) {
  params <- paraclu_params_twoargs(pos, score)

  if (!is.na(params$br)) {
    new_min  <- max(min_density, params$min_density)
    left     <-    1      : (params$br - 1)
    right    <- params$br :   length(pos)
    clusters <- rbind(paraclu_twoargs_df(pos[left],  score[left],  new_min, clusters), 
                      paraclu_twoargs_df(pos[right], score[right], new_min, clusters))
  }
  rbind( clusters
       , data.frame( start = min(pos)
                   ,   end = max(pos)
                   , min_d = min_density
                   , max_d = params$min_density))
}
paraclu_twoargs_df(l$pos, l$score) |> head()
```

Or do we lose time by keeping in memory slices of the original vector?

```{r paraclu_fourargs_df}
paraclu_fourargs_df <- function(pos, score, left, right, min_density = -Inf
                               , clusters = data.frame()) {
  range <- left : right
  params <- paraclu_params_twoargs(pos[range], score[range])
  br <- left - 1 + params$br

  if (!is.na(params$br)) {
    new_min     <- max(min_density, params$min_density)
    clusters <- rbind(paraclu_fourargs_df(pos, score, left, br - 1, new_min, clusters), 
                      paraclu_fourargs_df(pos, score, br  , right , new_min, clusters))
  }
  
  rbind( clusters
   , data.frame( start = min(pos)
            ,   end = max(pos)
            , min_d = min_density
            , max_d = params$min_density))
}
paraclu_fourargs_df(l$pos, l$score, 1, length(l$pos)) |> head()
```

## Results:

Using DataFrame or IRanges during recursion is very expensive, so let's not
bother benchmarking replicates.

```{r benchmarks_BioC}
microbenchmark::microbenchmark(times = 1
, tibble      = paraclu_twoargs_tibble(l$pos, l$score)
, data.frame  = paraclu_twoargs_df(l$pos, l$score)
, DataFrame   = paraclu_twoargs_DF(l$pos, l$score)
, IRanges     = paraclu_twoargs_IRanges(l$pos, l$score)
, IRanges_P   = paraclu_Pair_IRanges(pair)
, orig_simple = paraclu_orig_simpler_output(ctss.df)
, orig        = paraclu_orig(ctss.df)
)
```

How about the other methods ?

```{r benchmarks}
(benchmark <- microbenchmark::microbenchmark(times = 20
, tibble      = paraclu_twoargs_tibble(l$pos, l$score)
, data.frame  = paraclu_twoargs_df(l$pos, l$score)
, orig_simple = paraclu_orig_simpler_output(ctss.df)
, orig        = paraclu_orig(ctss.df)
, fourargs    = paraclu_fourargs_df(l$pos, l$score, 1, length(l$pos))
))

library("ggplot2") |> suppressPackageStartupMessages()
ggplot(benchmark, aes(x = time / 1e9, y = expr, color = expr)) +
  geom_boxplot() + 
  scale_x_log10("time (seconds)")
```

The `data.frame` method is simplest and fastest.

```{r result}
orig_df <- paraclu_orig(ctss.df)
new_df  <- paraclu_twoargs_df(l$pos, l$score)
all( identical(orig_df$start, new_df$start)
   , identical(orig_df$end  , new_df$end)
   , identical(orig_df$min_d, new_df$min_d)
   , identical(orig_df$max_d, new_df$max_d))
paraclu_twoargs_df
```

## Session information

```{r}
sessionInfo()
```