---
title: "rowsum on Rle DataFrames"
author: "Charles Plessy"
date: "2023-06-30"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

## Purpose

There is no native `rowsum` function for `DataFrame`s of `Rle`-encoded
numerical values.  What is the best strategy for an implementation ?

## Setup

```{r setup}
library("CAGEr")   |> suppressPackageStartupMessages()
library("ggplot2") |> suppressPackageStartupMessages()
```

The DataFrame encodes counts of CAGE tags mapped on the zebrafish genome.

```{r define_data}
# (DF <- CTSStagCountDF(exampleCAGEexp))
# DF[[1]]
# 
# (names <- CTSScoordinatesGR(exampleCAGEexp)$cluster)

(DF <- ZebrafishDevelopmentalCAGE::ZebrafishCAGE |> assay())
DF[[1]]
# Artificial example.
(names <- rep(CTSScoordinatesGR(exampleCAGEexp)$cluster, 711)[1:nrow(DF)])
```

The functions:

```{r define_functions}
.rowsumDelayedArrayAsMatrix <- function(x, group, reorder = TRUE)
  rowsum(
    as.matrix(DelayedArray::DelayedArray(x)),
    group,
    reorder = reorder) |> DataFrame()
.rowsumDelayedArrayAsMatrix(DF, decode(names), reorder = FALSE) |> head()

.rowsumDelayedArrayAsdataframe <- function(x, group, reorder = TRUE)
  rowsum(
    as.data.frame(DelayedArray::DelayedArray(x)),
    group,
    reorder = reorder) |> DataFrame()
.rowsumDelayedArrayAsdataframe(DF, decode(names), reorder = FALSE) |> head()

.rowsumDecodeAsDF <- function(x, group, reorder = TRUE)
  rowsum(
    as.data.frame(lapply(x, decode)),
    group,
    reorder = reorder) |> DataFrame()
.rowsumDecodeAsDF(DF, decode(names), reorder = FALSE) |> head()

# https://support.bioconductor.org/p/99960/#100034
.rowsum.splitAsList <- function(x, group, reorder=TRUE, ...)
    lapply(x, function(col) sum(splitAsList(col, group))) |> DataFrame()
.rowsum.splitAsList(DF, decode(names), reorder = FALSE) |> head()

.rowsum.decodeSplitAsList <- function(x, group, reorder=TRUE, ...)
    lapply(lapply(x, decode), function(col) sum(splitAsList(col, group))) |> DataFrame()
.rowsum.decodeSplitAsList(DF, decode(names), reorder = FALSE) |> head()

# This one I thought about after asking ChatGPT for a splitAsList alternative
.rowsum.splitAndColSums <- function(x, group, reorder=TRUE, ...) {
  l <- split(x, group)
  ll <- lapply(l, \(l) sapply(l, sum))
  do.call(rbind, ll) |> DataFrame()
}
.rowsum.splitAndColSums(DF, decode(names), reorder = FALSE) |> head()
.rowsum.splitAndColSums(endoapply(DF, decode), decode(names), reorder = FALSE) |> head()
```

## Benchmark

```{r benchmark, fig.height=3, fig.width=9}
(microbench_out <- microbenchmark::microbenchmark(
  times = 100,
  .rowsumDelayedArrayAsMatrix = .rowsumDelayedArrayAsMatrix(DF, decode(names), reorder = FALSE),
  .rowsumDelayedArrayAsdataframe = .rowsumDelayedArrayAsdataframe(DF, decode(names), reorder = FALSE),
  .rowsumDecodeAsDF = .rowsumDecodeAsDF(DF, decode(names), reorder = FALSE),
  .rowsum.splitAsList = .rowsum.splitAsList(DF, decode(names), reorder = FALSE),
  .rowsum.decodeSplitAsList = .rowsum.decodeSplitAsList(DF, decode(names), reorder = FALSE) #,
  # Interesting, # .rowsum.splitAndColSums = .rowsum.splitAndColSums(DF, decode(names), reorder = FALSE),
  # but too slow # .rowsum.DecodeSplitAndColSums = .rowsum.splitAndColSums(endoapply(DF, decode), decode(names), reorder = FALSE)
))
# https://statisticsglobe.com/microbenchmark-package-r
ggplot(microbench_out, aes(x = time / 1e9, y = expr, color = expr)) +  # Plot performance comparison
  geom_boxplot() + 
  scale_x_log10("time (seconds)")
```

## Result

The winner is:

```
.rowsumDecodeAsDF <- function(x, group, reorder = TRUE)
  rowsum(
    as.data.frame(lapply(x, decode)),
    group,
    reorder = reorder) |> DataFrame()
```

Needless to say, the `reorder` argument needs to be implemented with care.

## Session information

```{r sessionInfo}
sessionInfo()
```

## ChatGPT

Charles Plessy  7:52 PM

> Good evening ChatGPT, can you tell me an alternative to the Bioconductor function `splitAsList` ?

Sidekick APP  7:52 PM

> An alternative to the `splitAsList` function from Bioconductor in R could be the combination of the `split` and `as.list` functions from base R. Here's an example:

```
r
# Assuming 'df' is your data frame and 'group' is the column you want to split by
split_list = as.list(split(df, df$group))
```

> This will give you a list of data frames, split by the 'group' column, similar to what `splitAsList` would do.