---
title: "Fastest way to deduplicate pairs after reducing precision"
author: "Charles Plessy"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document: 
    keep_md: yes
---

To accelerate plotting and reduce the file size of vector graphics created with
the `plotCorrelation2` method, points are deduplicated after reducing their
precision to a smaller number of significant digits.  After benchmarking the
alternatives below, I chose the approach based on complex numbers, because their
performance is similar to `data.table` or `dplyr` objects, and because these
functions are available from standard R installations.

```{r setup, echo=FALSE}
options(width=120)
knitr::opts_chunk$set(cache  = TRUE, cache.lazy = FALSE)
knitr::opts_knit$set(verbose = TRUE)
```

```{r create_data}
set.seed(1)
pois1 <- rpois(1e7, 0.3)
pois2 <- rpois(1e7, 0.3)

nbin1 <- rnbinom(1e7, mu = 2, size = .1)
nbin2 <- rnbinom(1e7, mu = 2, size = .1)

# Same with a bit of noise
x     <- nbin1 * runif(1e7)
y     <- nbin2 * runif(1e7)

sx    <- signif(x, 2)
sy    <- signif(y, 2)
```

`DataFrame` tables are coerced to `data.frame` objects by the
`duplicated.DataFrame` function, therefore execution times are similar.

```{r declare_functions}
f.df       <- function(x,y) {
  df <- unique(data.frame(x = x, y = y))
  data.frame(df$x, df$y)}

f.tbl <- function(x,y) {
  tbl <- dplyr::distinct(dplyr::tibble(x = x, y = y))
  data.frame(tbl$x, tbl$y)}

f.dt       <- function(x,y) {
  dt <- unique(data.table::data.table(x = x, y = y))
  data.frame(dt$x, dt$y)}

f.cplx     <- function(x,y) {
  u <- unique(complex(real=x, im=y))
  data.frame(Re(u), Im(u))}

f.cplx.Rle <- function(x,y) {
  u <- unique(S4Vectors::Rle(complex(real=x, im=y)))
  data.frame(Re(u), Im(u))}
```


```{r benchmark_pois}
microbenchmark::microbenchmark(
  f.df       (pois1, pois2),
  f.dt       (pois1, pois2),
  f.tbl      (pois1, pois2),
  f.cplx     (pois1, pois2),
  f.cplx.Rle (pois1, pois2),  times = 10L)
```

```{r benchmark_nbin}
microbenchmark::microbenchmark(
  f.df       (nbin1, nbin2),
  f.dt       (nbin1, nbin2),
  f.tbl      (nbin1, nbin2),
  f.cplx     (nbin1, nbin2),
  f.cplx.Rle (nbin1, nbin2),  times = 10L)
```

```{r benchmark_xy}
microbenchmark::microbenchmark(
  f.df       (x, y),
  f.dt       (x, y),
  f.tbl      (x, y),
  f.cplx     (x, y),
  f.cplx.Rle (x, y),  times = 10L)
```

```{r benchmark_sxy}
microbenchmark::microbenchmark(
  f.df       (sx, sy),
  f.dt       (sx, sy),
  f.tbl      (sx, sy),
  f.cplx     (sx, sy),
  f.cplx.Rle (sx, sy),  times = 10L)
```
